### 2025-04-15 evaluating-prediction-accuracy.qmd
20250415 - R-squared and cross-refs

##### Mean Error {#sec-meanError}

@@

The [coefficient of determination](#sec-multipleRegressionRSquared) ($R^2$) reflects the proportion of variance in the outcome (dependent) variable that is explained by the model predictions: $R^2 = \frac{\text{variance explained in }Y}{\text{total variance in }Y}$.
Larger values indicate greater accuracy.

@@

Adjusted $R^2$ is similar to the [coefficient of determination](#sec-coefficientOfDetermination), but it accounts for the number of predictors included in the regression model to penalize [overfitting](#sec-overfitting).
Adjusted $R^2$ reflects the proportion of variance in the outcome (dependent) variable that is explained by the model predictions over and above what would be expected to be accounted for by chance, given the number of predictors in the model.
Larger values indicate greater accuracy.
The formula for adjusted $R^2$ is in @eq-adjustedRSquared.
Adjusted $R^2$ is described further in @sec-multipleRegressionRSquared.

>>>>

##### Mean Error (ME) {#sec-meanError}

@@

The [coefficient of determination](#sec-multipleRegressionRSquared) ($R^2$) is described in @sec-multipleRegressionRSquared.
The [coefficient of determination](#sec-multipleRegressionRSquared) reflects the proportion of variance in the outcome (dependent) variable that is explained by the model predictions: $R^2 = \frac{\text{variance explained in }Y}{\text{total variance in }Y}$.
Larger values indicate greater accuracy.

@@

Adjusted $R^2$ ($R^2_{adj}$) is described in @sec-multipleRegressionAdjustedRsquared.
Adjusted $R^2$ is similar to the [coefficient of determination](#sec-coefficientOfDetermination), but it accounts for the number of predictors included in the regression model to penalize [overfitting](#sec-overfitting).
Adjusted $R^2$ reflects the proportion of variance in the outcome (dependent) variable that is explained by the model predictions over and above what would be expected to be accounted for by chance, given the number of predictors in the model.
Larger values indicate greater accuracy.
The formula for adjusted $R^2$ is in @eq-adjustedRSquared.

### 2025-04-15 multiple-regression.qmd
20250415 - R-squared and cross-refs

Regression with multiple predictors—i.e., multiple regression—takes the form of @eq-multipleRegression:

@@

- there is a linear association between the predictor variables and the outcome variable
- there is homoscedasticity of the residuals; the residuals do not differ as a function of the predictor variables or as a function of the outcome variable
- the residuals are independent; they are uncorrelated with each other
- the residuals are normally distributed

Homoscedasticity of the residuals means that the variance of the residuals does not differ as a function of the outcome variable or as a function of the predictor variables (i.e., the residuals show constant variance as a function of outcome/predictors).

@@

To evaluate the shape of the association between the predictor variables and the outcome variable, we can examine scatterplots, residual plots, marginal model plots, and added-variable plots.

@@

## Coefficient of Determination ($R^2$) {#sec-multipleRegressionRSquared}

@@

One issue with $R^2$ is that it increases as the number of predictors increases, which can lead to [overfitting](#sec-overfitting) if using $R^2$ as an index to compare models for purposes of selecting the "best-fitting" model.
Consider the following example (adapted from @PetersenPrinciplesPsychAssessment) in which you have one [predictor variable](#sec-correlationalStudy) and one [outcome variable](#sec-correlationalStudy), as shown in @tbl-regression1.

@@

Using the data, the best fitting regression model is: $y =$ `{r} regression1_intercept` $+$ `{r} regression1_slope` $\cdot x_1$.
In this example, the $R^2$ is `{r} regression1_rsquare`.
The equation is not a perfect prediction, but with a single [predictor variable](#sec-correlationalStudy), it captures the majority of the variance in the outcome.

@@

With the second [predictor variable](#sec-correlationalStudy), the best fitting regression model is: $y =$ `{r} regression2_intercept` + `{r} regression2_slope1` $\cdot x_1 +$ `{r} regression2_slope2` $\cdot x_2$.
In this example, the $R^2$ is `{r} regression2_rsquare`.
The equation with the second [predictor variable](#sec-correlationalStudy) provides a perfect prediction of the outcome.

@@

As an example, consider that we have 13 [predictor variables](#sec-correlationalStudy) to predict fantasy performance for 43 players.
Assume that, with 13 [predictor variables](#sec-correlationalStudy), we explain 38% of the variance ($R^2 = .38; r = .62$).
We explained a lot of the variance in the outcome, but it is important to consider how much variance could have been explained by random chance: $E(R^2) = \frac{K}{n-1} = \frac{13}{43 - 1} = .31$.
We expect to explain 31% of the variance, by chance, in the outcome.
So, 82% of the variance explained was likely spurious (i.e., $\frac{.31}{.38} = .82$).
As the sample size increases, the spuriousness decreases.

To account for the number of [predictor variables](#sec-correlationalStudy) in the model, we can use a modified version of $R^2$ called adjusted $R^2$ ($R^2_{adj}$).
Adjusted $R^2$ ($R^2_{adj}$) accounts for the number of [predictor variables](#sec-correlationalStudy) in the model, based on how much would be expected to be accounted for by chance to penalize [overfitting](#sec-overfitting).
Adjusted $R^2$ ($R^2_{adj}$) reflects the proportion of variance in the [outcome (dependent) variable](#sec-correlationalStudy) that is explained by the model predictions over and above what would be expected to be accounted for by chance, given the number of [predictor variables](#sec-correlationalStudy) in the model.
The formula for adjusted $R^2$ ($R^2_{adj}$) is in @eq-adjustedRSquared:

@@

Let's say we want to use a number of variables to predict a wide receiver's fantasy performance.
We want to consider several predictors, including the player's age, height, weight, and target share.
We have only a few predictors and our sample size is large enough such that [overfitting](#sec-overfitting) is not likely a concern.

@@

There are some suggestions of potential nonlinearity, such as an inverted-U-shaped association between height and fantasy points, suggesting that there may an optimal range for height among Wide Receivers—being too short or too tall could be a disadvantage.
In addition, target share shows a weakening association as target share increases.
Thus, after evaluating the linear association between the predictors and outcome, we will also examine the possibility for curvilinear associations.

@@

The only terms that are significantly associated with fantasy performance among Wide Receivers are weight and target share, both of which showed a positive association with fantasy points.
If we want to obtain standardized regression coefficients, we can standardize the outcome variable and each predictor variable using the `scale()` function:

@@

Target share has a large [effect size](#sec-practicalSignificance).
All of the other predictors have small effect sizes.

@@

### Evaluating and Addressing Assumptions {#sec-multipleRegressionExamplesAddressingAssumptions}

@@

- there is a linear association between the predictor variables and the outcome variable
- there is homoscedasticity of the residuals; the residuals do not differ as a function of the predictor variables or as a function of the outcome variable
- the residuals are independent; they are uncorrelated with each other
- the residuals are normally distributed

#### Linear Association {#sec-multipleRegressionExamplesAddressingAssumptionsLinearAssociation}

We evaluated the shape of the association between the predictor variables and the outcome variables using scatterplots.
We accounted for potential curvilinearity with a quadratic term.
Other ways to account for nonlinearity, in addition to polynomials (e.g., quadratic, cubic, quartic, or higher-degree terms), include transforming predictors (e.g., log, square root, inverse, exponential transform), splines/piecewise regression, and generalized additive models.

@@

#### Homoscedasticity {#sec-multipleRegressionExamplesAddressingAssumptionsHomoscedasticity}

@@

In this example, the residuals appear to increase as a function of the fitted values.
To handle this, we may need to transform the outcome variable to be more normally distributed.

@@

The residuals show more constant variance after transforming the outcome variable.

#### Uncorrelated Residuals {#sec-multipleRegressionExamplesAddressingAssumptionsUncorrelatedResiduals}

@@

#### Normally Distributed Residuals {#sec-multipleRegressionExamplesAddressingAssumptionsNormallyDistributedResiduals}

@@

To address multicollinearity, you can drop a redundant predictor or you can also use principal component analysis or factor analysis of the predictors to reduce the predictors down to a smaller number of meaningful predictors.
For a meaningful answer in a regression framework that is precise and confident, you need a low level of intercorrelation among predictors, unless you have a very large sample size.

@@

- columns with non-zero values are predictors of the variable specified in the given row
- the diagonal of the predictor matrix should be zero because a variable cannot predict itself

@@

- When computing the interaction term, first mean-center the predictor variables.
Calculate the interaction term as the multiplication of the mean-centered predictor variables.
Mean-centering the predictor variables when computing the interaction term is important for addressing issues regarding [multicollinearity](#sec-multipleRegressionMulticollinearity) [@Iacobucci2016].
- When including an interaction term in the model, make sure also to include the main effects.

First, we mean-center the predictors.
In this case, we center the predictors around the mean of height and weight for Wide Receivers:

@@

Then, we compute the interaction term as the multiplication of the two centered predictors:

@@

Multiple regression allows examining the association between multiple [predictor variables](#sec-correlationalStudy) and one [outcome variable](#sec-correlationalStudy).
Inclusion of multiple predictors in the model allows for potentially greater predictive accuracy and identification of the extent to which each variable uniquely contributes to the outcome variable.
As with [correlation](#sec-correlation), an association does not imply causation.
However, identifying associations is important because associations are a necessary (but insufficient) condition for causality.
When developing a multiple regression model, it is important to pay attention for potential [multicollinearity](#sec-multipleRegressionMulticollinearity)—it may become difficult to detect a given [predictor variable](#sec-correlationalStudy) as [statistically significant](#sec-statisticalSignificance) due to the greater uncertainty around the parameter estimates.

>>>>

Regression with multiple [predictors](#sec-correlationalStudy)—i.e., multiple regression—takes the form of @eq-multipleRegression:

@@

- there is a linear association between the [predictor variable](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy)
- there is homoscedasticity of the residuals; the residuals do not differ as a function of the [predictor variable](#sec-correlationalStudy) or as a function of the [outcome variable](#sec-correlationalStudy)
- the residuals are independent; they are uncorrelated with each other
- the residuals are normally distributed

Homoscedasticity of the residuals means that the variance of the residuals does not differ as a function of the [outcome variable](#sec-correlationalStudy) or as a function of the [predictor variable](#sec-correlationalStudy) (i.e., the residuals show constant variance as a function of [outcome](#sec-correlationalStudy)/[predictors](#sec-correlationalStudy)).

Those are some of the key assumptions of multiple regression.
However, there are additional assumptions of multiple regression, including ones discussed in the chapter on [Causal Inference](#sec-causalInference).
For instance, the variables included should reflect causal process such that the [predictor variables](#sec-correlationalStudy) influence the [outcome variable](#sec-correlationalStudy), and not the other way around.
That is, the [outcome variable](#sec-correlationalStudy) should not influence the [predictor variables](#sec-correlationalStudy) (i.e., there should be no reverse causation).
In addition, it is important control for any [confound(s)](#sec-causalDiagramConfounding).
If a [confound](#sec-causalDiagramConfounding) is not controlled for, this is called omitted-variable bias, and it leads the researcher to incorrectly attribute the effects of the omitted variable to the included variables.

@@

To evaluate the shape of the association between the [predictor variables](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy), we can examine scatterplots, residual plots, marginal model plots, and added-variable plots.

@@

## How Much Variance the Model Explains {#sec-multipleRegressionExplainedVariance}

When estimating a multiple regression model, it can be useful to evaluate how much variance in the [outcome variable](#sec-correlationalStudy) that the [predictor variable(s)](#sec-correlationalStudy) explain (i.e., account for).
If the variables collectively explain only a small amount of variance, it suggest that the [predictor variables](#sec-correlationalStudy) have a small [effect size](#sec-practicalSignificance) and that other [predictor variables](#sec-correlationalStudy) will be necessary to account for the majority of variability in the [outcome variable](#sec-correlationalStudy).
There are two primary indices of how much how much variance in the [outcome variable](#sec-correlationalStudy) is explained by the [predictor variable(s)](#sec-correlationalStudy): the [coefficient of determination](#sec-multipleRegressionRSquared) ($R^2$) and [adjusted $R^2$](#sec-multipleRegressionAdjustedRsquared) ($R^2_{adj}$), described below.

### Coefficient of Determination ($R^2$) {#sec-multipleRegressionRSquared}

@@

One issue with $R^2$ is that it increases as the number of [predictors](#sec-correlationalStudy) increases, which can lead to [overfitting](#sec-overfitting) if using $R^2$ as an index to compare models for purposes of selecting the "best-fitting" model.
Consider the following example (adapted from @PetersenPrinciplesPsychAssessment) in which you have one [predictor variable](#sec-correlationalStudy) and one [outcome variable](#sec-correlationalStudy), as shown in @tbl-regression1.

@@

Using the data, the best fitting regression model is: $y =$ `{r} regression1_intercept` $+$ `{r} regression1_slope` $\cdot x_1$.
In this example, the $R^2$ is `{r} regression1_rsquare`.
The equation is not a perfect prediction, but with a single [predictor variable](#sec-correlationalStudy), it captures the majority of the variance in the [outcome](#sec-correlationalStudy).

@@

With the second [predictor variable](#sec-correlationalStudy), the best fitting regression model is: $y =$ `{r} regression2_intercept` + `{r} regression2_slope1` $\cdot x_1 +$ `{r} regression2_slope2` $\cdot x_2$.
In this example, the $R^2$ is `{r} regression2_rsquare`.
The equation with the second [predictor variable](#sec-correlationalStudy) provides a perfect prediction of the [outcome](#sec-correlationalStudy).

@@

As an example, consider that we have 13 [predictor variables](#sec-correlationalStudy) to predict fantasy performance for 43 players.
Assume that, with 13 [predictor variables](#sec-correlationalStudy), we explain 38% of the variance ($R^2 = .38; r = .62$).
We explained a lot of the variance in the [outcome](#sec-correlationalStudy), but it is important to consider how much variance could have been explained by random chance: $E(R^2) = \frac{K}{n-1} = \frac{13}{43 - 1} = .31$.
We expect to explain 31% of the variance, by chance, in the [outcome](#sec-correlationalStudy).
So, 82% of the variance explained was likely spurious (i.e., $\frac{.31}{.38} = .82$).
As the sample size increases, the spuriousness decreases.
To account for the number of [predictor variables](#sec-correlationalStudy) in the model, we can use a modified version of $R^2$ called adjusted $R^2$ ($R^2_{adj}$), described next.

### Adjusted $R^2$ ($R^2_{adj}$) {#sec-multipleRegressionAdjustedRsquared}

Adjusted $R^2$ ($R^2_{adj}$) accounts for the number of [predictor variables](#sec-correlationalStudy) in the model, based on how much would be expected to be accounted for by chance to penalize [overfitting](#sec-overfitting).
Adjusted $R^2$ ($R^2_{adj}$) reflects the proportion of variance in the [outcome (dependent) variable](#sec-correlationalStudy) that is explained by the model predictions over and above what would be expected to be accounted for by chance, given the number of [predictor variables](#sec-correlationalStudy) in the model.
The formula for adjusted $R^2$ ($R^2_{adj}$) is in @eq-adjustedRSquared:

@@

### Predicting Wide Receivers' Fantasy Points {#sec-multipleRegressionFantasyPointsWRs}

Let's say we want to use a number of variables to predict a wide receiver's fantasy performance.
We want to consider several [predictors](#sec-correlationalStudy), including the player's age, height, weight, and target share.
We have only a few [predictors](#sec-correlationalStudy) and our sample size is large enough such that [overfitting](#sec-overfitting) is not likely a concern.

#### Examine Bivariate Associations {#sec-multipleRegressionFantasyPointsWRsBivariate}

@@

There are some suggestions of potential nonlinearity, such as an inverted-U-shaped association between height and fantasy points, suggesting that there may an optimal range for height among Wide Receivers—being too short or too tall could be a disadvantage.
In addition, target share shows a weakening association as target share increases.
Thus, after evaluating the linear association between the [predictors](#sec-correlationalStudy) and [outcome](#sec-correlationalStudy), we will also examine the possibility for curvilinear associations.

#### Estimate Multiple Regression Model {#sec-multipleRegressionFantasyPointsWRsModel}

@@

The only terms that are significantly associated with fantasy performance among Wide Receivers are weight and target share, both of which showed a positive association with fantasy points.

We can obtain the [coefficient of determination](#sec-multipleRegressionRSquared) ($R^2$) and [adjusted $R^2$](#sec-multipleRegressionAdjustedRsquared) ($R^2_{adj}$) using the following code:

```{r}
summary(linearRegressionModel)$r.squared
summary(linearRegressionModel)$adj.r.squared
```

```{r}
#| include: false

linearRegressionModel_rsquared <- summary(linearRegressionModel)$r.squared
```

The model explained `r petersenlab::apa(linearRegressionModel_rsquared * 100, decimals = 0)`% of the variability in fantasy points (i.e., $R^2 = `r petersenlab::apa(linearRegressionModel_rsquared, decimals = 2, leading = FALSE)`$).

If we want to obtain standardized regression coefficients, we can standardize the [outcome variable](#sec-correlationalStudy) and each [predictor variable](#sec-correlationalStudy) using the `scale()` function:

@@

Target share has a large [effect size](#sec-practicalSignificance).
All of the other [predictors](#sec-correlationalStudy) have small [effect size](#sec-practicalSignificance).

@@

#### Evaluating and Addressing Assumptions {#sec-multipleRegressionExamplesAddressingAssumptions}

@@

- there is a linear association between the [predictor variables](#sec-correlationalStudy) and the [outcome variable](#sec-correlationalStudy)
- there is homoscedasticity of the residuals; the residuals do not differ as a function of the [predictor variables](#sec-correlationalStudy) or as a function of the [outcome variable](#sec-correlationalStudy)
- the residuals are independent; they are uncorrelated with each other
- the residuals are normally distributed

##### Linear Association {#sec-multipleRegressionExamplesAddressingAssumptionsLinearAssociation}

We evaluated the shape of the association between the [predictor variables](#sec-correlationalStudy) and the [outcome variables](#sec-correlationalStudy) using scatterplots.
We accounted for potential curvilinearity with a quadratic term.
Other ways to account for nonlinearity, in addition to polynomials (e.g., quadratic, cubic, quartic, or higher-degree terms), include transforming [predictors](#sec-correlationalStudy) (e.g., log, square root, inverse, exponential transform), splines/piecewise regression, and generalized additive models.

@@

##### Homoscedasticity {#sec-multipleRegressionExamplesAddressingAssumptionsHomoscedasticity}

@@

In this example, the residuals appear to increase as a function of the fitted values.
To handle this, we may need to transform the [outcome variable](#sec-correlationalStudy) to be more normally distributed.

@@

The residuals show more constant variance after transforming the [outcome variable](#sec-correlationalStudy).

##### Uncorrelated Residuals {#sec-multipleRegressionExamplesAddressingAssumptionsUncorrelatedResiduals}

@@

##### Normally Distributed Residuals {#sec-multipleRegressionExamplesAddressingAssumptionsNormallyDistributedResiduals}

@@

To address multicollinearity, you can drop a redundant predictor or you can also use [principal component analysis](#sec-pca) or [factor analysis](#sec-factorAnalysis) of the [predictors](#sec-correlationalStudy) to reduce the [predictors](#sec-correlationalStudy) down to a smaller number of meaningful [predictors](#sec-correlationalStudy).
For a meaningful answer in a regression framework that is precise and confident, you need a low level of intercorrelation among [predictors](#sec-correlationalStudy), unless you have a very large sample size.

@@

- columns with non-zero values are [predictors](#sec-correlationalStudy) of the variable specified in the given row
- the diagonal of the predictor matrix should be zero because a variable cannot predict itself

@@

- When computing the interaction term, first mean-center the [predictor variables](#sec-correlationalStudy).
Calculate the interaction term as the multiplication of the mean-centered [predictor variables](#sec-correlationalStudy).
Mean-centering the [predictor variables](#sec-correlationalStudy) when computing the interaction term is important for addressing issues regarding [multicollinearity](#sec-multipleRegressionMulticollinearity) [@Iacobucci2016].
- When including an interaction term in the model, make sure also to include the main effects.

First, we mean-center the [predictors](#sec-correlationalStudy).
In this case, we center the [predictors](#sec-correlationalStudy) around the mean of height and weight for Wide Receivers:

@@

Then, we compute the interaction term as the multiplication of the two centered [predictors](#sec-correlationalStudy):

@@

Multiple regression allows examining the association between multiple [predictor variables](#sec-correlationalStudy) and one [outcome variable](#sec-correlationalStudy).
Inclusion of multiple [predictors](#sec-correlationalStudy) in the model allows for potentially greater predictive accuracy and identification of the extent to which each variable uniquely contributes to the [outcome variable](#sec-correlationalStudy).
As with [correlation](#sec-correlation), an association does not imply causation.
However, identifying associations is important because associations are a necessary (but insufficient) condition for causality.
When developing a multiple regression model, it is important to pay attention for potential [multicollinearity](#sec-multipleRegressionMulticollinearity)—it may become difficult to detect a given [predictor variable](#sec-correlationalStudy) as [statistically significant](#sec-statisticalSignificance) due to the greater uncertainty around the parameter estimates.
